---
title: "PS270 Final Project"
author: "Luca Stewart-Mariuci"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

#### Schools Districts Spending: What Effect Does School District Spending Have On Quality and Performance of Education

### Introduction

For this final project I wanted to look at the relationship of the amount of money a school district spends per student and the school district performance. This is done by looking at the impact the Per-Pupil Price has on the Overall Accountability Score. Wisconsin calculates the cost of education in three different ways, because of that we will look at all three to better understand the impact that school funding has on the education outcomes. These three calculations of school districts spending are `Total Current Education Cost,` `Total Education Cost,` and `Total District Cost.`
  If the a school district has a higher Per-Pupil Price, in any one of the three calculation methods, then the Overall Accountability Score will be higher for that school district. This is because the more money a school district spends on an individual student, then in turn that student would have more support and have higher scores in both academics and personal growth. 
  
This question and research topic is important because funding is one the go-to ways in conversations about how to improve schools. If data suggests that funding in education is has positive effects on school performance then increasing funding could be beneficial, or if it has a neutral or negative effect then it would suggest that their are other factors in play that have a large impact on the quality of education.

### Data Section

# looking at the variables

Overall Accountability Score
  This score is based of the performance of students in a public school district.   There are multiple factors in the calculation these include performance in        subjects, growth of students in different subjects, the ability to close the      achievement gaps, students determined to be on track, ability to help English     Second Language students, students in special education programs, and post        secondary readiness of students. 
    This score is on 0.0 - 100.00 point scale, with 0.0 being the worst a school    distinct can do, and 100.00 being the best. Then it assigns a report card score   which consists of
    Fails to Meet Expectations: which 0.0 - 48.9
    Meets Few Expectations: 49.0 - 59.9
    Meets Expectations: 60.0 - 70.9
    Exceeds Expectations: 71.0 - 83.9
    Significantly Exceeds Expectations: 84.0 - 100
  (zero schools in Wisconsin Fail to Meet Expectations)

Per-Pupil Price
  This score is the average amount of spending divided by the total member/student   count.
    The Wisconsin Department of Public Instruction has three ways to calculate the   total costs. Because of this, and wanting to ensure that there are no             discrepancies between the different ways to calculate the cost of education this   project will look at all three. 
    Total Current Education Costs (TCEC): Includes the general fund instructional costs, Special project fund instructional costs, and instructional support services
    Total Education Costs (TEC): TCEC + transportation, facility acquisition costs, and debt services.
    Total District Costs (TDC): Total district spending, TCEC + TEC and things like food and community based services. 

  Then by dividing the total cost per each total by member/student we get the Per-Pupil Price of all three total sums TCEC Per-Pupil Price, TEC Per-Pupil Price, and TDC Per-Pupil Price.
  
# Loading Pachages and Data

```{r}
# Loading needed packages into R.
library(tidyverse)
library(dplyr)
library(ggplot2)
library(broom)
library(modelsummary)
```

```{r}
# Loading Report Card scores into R.
library(readxl)
X2023_2024_district_report_card <- read_excel("2023-2024_district_report_card.xlsx")

# This will give us the data for the Overall Accountability Score.
```
```{r}
# Loading school districts expenditures into R.
library(readxl)
cmpcst24_1_ <- read_excel("cmpcst24 (1).xlsx")

# This will give us the data for the different education costs totals and the total number of students per district.
```

### Final Project Code

# Selecting Data Needed Cleaning Up The Data Sets

With the importing of the data there were some issues of how the data was displayed and how it is interpreted, this section cleans it up and fixes all that.

```{r}
# Reassigning the data and selecting what I need.
report_card <- X2023_2024_district_report_card 

report_card <- report_card |>
  select("District Name", "Overall Accountability Score", "Overall Accountability Rating",
         "Grade Band for Comparison", "District Enrollment") 
```

```{r}
# Reassigning the data and selecting what I need.
district_cost <- cmpcst24_1_

district_cost <- district_cost |>
  select("...5", "...6", "...9", "...11")
```

```{r}
#Giving proper names to the columns.
district_cost <- district_cost |>
  rename(
    `District Name` = `...5`,
    `TCEC` = `...6`,
    `TEC` = `...9`,
    `TDC` = `...11`) 
```

```{r}
# Getting rid the errors in the importing of the education cost data set.
district_cost <- district_cost |>
  filter(
  `District Name` != " ", 
  `District Name` != "NAME", 
  `District Name` != "18310.044339754681",
  `District Name` != "Statewide Total")
  
```

Joining The Data Sets

```{r}
# joining the report card score to the district score.
school_district <- district_cost |>
  inner_join(report_card)

# Joining the data on the school district name and then creating a new variable with all of the data.
```

Creating and Editing Variables

```{r}
# Fixes the variables so R reads them as numeric values.
school_district <- school_district |>
  mutate(
    "TCEC" = as.numeric(`TCEC`),
    "TEC" = as.numeric(`TEC`),
    "TDC" = as.numeric(`TDC`),
    "Overall Accountability Score" = as.numeric(`Overall Accountability Score`))
```

```{r}
# Creating-the-Per-Pupil-Price
school_district <- school_district |>
    mutate(
      "TCEC Per-Pupil Price" = `TCEC` / `District Enrollment`,
      "TEC Per-Pupil Price" = `TEC` / `District Enrollment`,
      "TDC Per-Pupil Price" = `TDC` / `District Enrollment`
      )
# This calculates all of the Per-Pupil Prices.
```

# Looking at the school_district and visualizing it

To better visualize the data here is a Bar Plot, this plot showcases how many schools fall into each report card score category.

```{r plot-only, echo=FALSE}
# Making a graph that showcases the amount of schools that fall in each report report card score
report_card_bar_chart <- school_district |>
   mutate(`Overall Accountability Rating` = factor(
    `Overall Accountability Rating`,
    levels = c(
      "Meets Few Expectations",
      "Meets Expectations",
      "Exceeds Expectations",
      "Significantly Exceeds Expectations"
    ) # This changes the order in which they apear on the graph
  )) |>
  ggplot(aes(x = `Overall Accountability Rating`)) +
  geom_bar(fill = "steelblue", color = "black", ) +
  labs(
    title = "Distribution of Accountability Ratings Across Wisconsin Districts",
    x = "Overall Accountability Rating",
    y = "Number of Districts",
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1)) 
# Changes angle of text to prevent overlap.

report_card_bar_chart
```


## Looking at diffrent means of calculating the price

# Looking at TCEC

```{r plot-only, echo=FALSE}
# Making the graph for the distribution of TCEC
tcec_school_district_graph <- school_district |>
  ggplot(aes(x = `TCEC Per-Pupil Price`,
             y = `Overall Accountability Score`)) + 
  geom_point() +
  labs(
    title = "Impact of The Total Current Education Cost Per Student and Impact it Has",
    subtitle = "Comparison of The TCEC Per-Pupil Price and The Overall Accountavility Score Per School Distirct",
    x = "TCEC Per-Pupil Price Per Distirct",
    y = "Overall Accountability Score Per District",
    caption = "Wisconsin Department of Public Instruction"
  )
tcec_school_district_graph
```


# Looking at TEC

```{r plot-only, echo=FALSE}
tec_school_district_graph <- school_district |>
  ggplot(aes(x = `TCEC Per-Pupil Price`,
             y = `Overall Accountability Score`)) + 
  geom_point() +
  labs(
    title = "Impact of The Total Education Cost Per Student and Impact it Has",
    subtitle = "Comparison of The TEC Per-Pupil Price and The Overall Accountavility Score Per School Distirct",
    x = "TEC Per-Pupil Price Per Distirct",
    y = "Overall Accountability Score Per District",
    caption = "Wisconsin Department of Public Instruction"
  )

tec_school_district_graph
```

# Looking at TDC

```{r plot-only, echo=FALSE}
tdc_school_district_graph <- school_district |>
  ggplot(aes(x = `TDC Per-Pupil Price`,
             y = `Overall Accountability Score`)) + 
  geom_point() +
  labs(
    title = "Impact of The Total District Cost Per Student and Impact it Has",
    subtitle = "Comparison of The TDC Per-Pupil Price and The Overall Accountavility Score Per School Distirct",
    x = "TDC Per-Pupil Price Per Distirct",
    y = "Overall Accountability Score Per District",
    caption = "Wisconsin Department of Public Instruction"
  )

tdc_school_district_graph
```

```{r plot-only, echo=FALSE}
# Creating a Table with the mean of spending with each form of education cost based on the category 

mean_table <- school_district |>
  mutate(`Overall Accountability Rating` = factor(
    `Overall Accountability Rating`,
    levels = c(
      "Meets Few Expectations",
      "Meets Expectations",
      "Exceeds Expectations",
      "Significantly Exceeds Expectations"
    )
  )) |>
  group_by(`Overall Accountability Rating`) |>
  summarize(
    `Mean of Per-Pupil TCEC` = mean(`TCEC Per-Pupil Price`, na.rm = TRUE),
    `Mean of Per-Pupil TEC`  = mean(`TEC Per-Pupil Price`, na.rm = TRUE),
    `Mean of Per-Pupil TDC`  = mean(`TDC Per-Pupil Price`, na.rm = TRUE),
    Number = n()
  ) |>
  knitr::kable(digits = 2)

mean_table
```

# Regression

```{r plot-only, echo=FALSE}
# Graphing the regression of TCEC.
tcec_reg_school_district_graph <- tcec_school_district_graph +
  geom_smooth(method = "lm", color = "brown4", size = 1.5)

tcec_reg_school_district_graph
```

```{r}
# Finding the Regression of TCEC
tcec_reg <- lm(`Overall Accountability Score` ~ `TCEC Per-Pupil Price`, data = school_district)

tcec_reg |> summary()
 
```

```{r plot-only, echo=FALSE}
# Creating a Table of TCEC Regression
tcec_reg_table <- tcec_reg |> 
  broom::tidy() |> 
  kable(
    caption = "Regression of Accountability Score on Per-Pupil TCEC",
    col.names = c("Term", "Estimate", "Std. Error", "T-Stat", "P-Value"), 
    digits = 2
  )

```

```{r plot-only, echo=FALSE}
# Graphing the regression of TEC.
tec_reg_school_district_graph <- tec_school_district_graph +
  geom_smooth(method = "lm", color = "yellow1", size = 1.5)

tec_reg_school_district_graph
```
```{r}
# Finding the Regression of TEC
tec_reg <- lm(`Overall Accountability Score` ~ `TEC Per-Pupil Price`, data = school_district)

tec_reg |> summary()
 
```

```{r plot-only, echo=FALSE}
# Creating a Table of TEC Regression
tec_reg_table <- tec_reg |> 
  broom::tidy() |> 
  kable(
    caption = "Regression of Accountability Score on Per-Pupil TEC",
    col.names = c("Term", "Estimate", "Std. Error", "T-Stat", "P-Value"), 
    digits = 2
  )

```


```{r plot-only, echo=FALSE}
# Graphing the regression of TDC.
tdc_reg_school_district_graph <- tdc_school_district_graph +
  geom_smooth(method = "lm", color = "blue1", size = 1.5)

tdc_reg_school_district_graph
```

```{r}
# Finding the Regression of TDC
tdc_reg <- lm(`Overall Accountability Score` ~ `TDC Per-Pupil Price`, data = school_district)

tdc_reg |> summary()
 
```

```{r plot-only, echo=FALSE}
# Creating a Table of TDC Regression
tdc_reg_table <- tdc_reg |> 
  broom::tidy() |> 
  kable(
    caption = "Regression of Accountability Score on Per-Pupil TDC",
    col.names = c("Term", "Estimate", "Std. Error", "T-Stat", "P-Value"), 
    digits = 2
  )

```

When looking at the regression we can see that for each dollar spent there is 0.0007937 decrease in the Overall Accountability Score, in a larger scale for each 1000 dollars spent WI schools get 0.7937 percent worse. The P-Value is below 0.5 so their is a statistic correlation.

While there is a correlation between negative correlation between the funding of schools and their report card score, there are more factors in play then just funding for a schools performance.

```{r}
school_district
```

